\section{Conclusion}
\label{sec:conclusion}

The field of one-step generative modeling has matured rapidly, evolving from the foundational framework of flow matching to a rich ecosystem of theoretically grounded, empirically validated methods. We have traced this progression through three key velocity-based methodological advances: consistency flow matching, which enforces straightness through velocity consistency; mean flows, which enable training from scratch via the average velocity identity; and AlphaFlow, which unifies trajectory flow matching and MeanFlow through a curriculum over the consistency step ratio. Together, these methods represent a shift from simulation-based sampling to direct prediction of the noise-to-data transport.

Looking forward, several directions remain open. First, while current methods excel on image generation, extending them to high-dimensional modalities such as video, audio, and 3D remains challenging. Second, the interplay between one-step generation and controllability---whether through classifiers, text prompts, or reward functions---presents both opportunities and challenges. Third, understanding when training from scratch (MeanFlow, AlphaFlow) outperforms distillation or constrained training (Consistency-FM), and vice versa, is still an active area of research.

Ultimately, the goal of one-step flow matching is not merely to accelerate sampling, but to retain the full generative power of diffusion and flow models while removing the computational barrier to real-world deployment. The methods surveyed here---spanning from self-consistency bootstrapping (shortcut models) to analytic JVP-based identities (MeanFlow) and curriculum-driven unification (AlphaFlow)---suggest that this goal is within reach. Finally, while the focus of this survey is on computer vision, the extent to which these velocity-based one-step techniques transfer to other modalities---such as speech and audio generation---remains largely unexplored and warrants further investigation.
